---
title: "MATH 216 Homework 1"
author: "Shannia Fu"
output:
  pdf_document:
    toc: yes
  html_document:
    collapsed: no
    smooth_scroll: no
    toc: yes
    toc_float: yes
---

```{r, echo=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
# For read_csv() command, which is an improved version of base R's read.csv()
library(readr) 
library(lubridate)

# Load data sets. Note this assumes this .Rmd files is in the same directory as
# the .csv files.
flights <- read_csv("data/flights.csv") %>% 
  mutate(date=as.Date(date))
weather <- read_csv("data/weather.csv") %>% 
  mutate(date=as.Date(date))
planes <- read_csv("data/planes.csv")
airports <- read_csv("data/airports.csv")
states <- read_csv("data/states.csv")
```





## Admistrative:

Please indicate

* Who you collaborated with: Brenda Li
* Roughly how much time you spent on this HW so far: 5 hours
* The URL of the RPubs published URL [here](http://rpubs.com/shannia/hw1).
* What gave you the most trouble: #3
* Any comments you have:





## Question 1:

Plot a "time series" of the proportion of flights that were delayed by > 30 minutes on each day.  i.e.
 
* the x-axis should be some notion of time
* the y-axis should be the proportion.


Using this plot, indicate describe the
[seasonality](https://en.wikipedia.org/wiki/Seasonality) of when delays over 30
minutes tend to occur.




```{r, echo=FALSE, message=FALSE, fig.width=12, fig.height=6}
all_flights <- flights %>%
  group_by(date) %>% 
  filter(!is.na(dep_delay)) %>% 
  mutate(over30=(ifelse(dep_delay>30, TRUE, FALSE))) %>% 
  group_by(date) %>% 
  summarise(prop=mean(over30))
  #View()
#all_flights

flight_time_series <- ggplot(all_flights, aes(x = date, y = prop)) + labs(title="Proportion of delayed flights daily from Jan 2011 to Jan 2012", y="proportion of delayed flights per day") +
  geom_point()
flight_time_series
```
From the graph, it seems that delays typically occur more often in July-August (summer), and in December-January (winter). There's also a lower portion of delays just around October (fall).

## Question 2:

Some people prefer flying on older planes.  Even though they aren't as nice,
they tend to have more room.  Which airlines should these people favor?


```{r, echo=FALSE, fig.width=12, fig.height=6}
older_planes <-
  inner_join(flights, planes, by="plane")
older_planes %>% 
  filter(!is.na(year)) %>% 
  filter(year < 1995) %>% 
  arrange(year) %>% 
  group_by(carrier) %>% 
  tally() %>% 
  arrange(desc(n)) %>% 
  knitr::kable()

older_planes %>% 
  filter(!is.na(year)) %>% 
  mutate(older = year <1995) %>% 
  group_by(carrier) %>% 
  summarise(p=mean(older)) %>%
  filter(p>0) %>% 
  mutate(p=round(p, 3)) %>% 
  arrange(desc(p)) %>% 
  knitr::kable()

older_planes %>% 
  filter(!is.na(year)) %>% 
  mutate(older = ifelse(year <1995, "from before 1995", "from after 1995")) %>% 
  group_by(carrier) %>% 
  count(older) %>% 
  ggplot(aes(x=carrier,y=n, fill=older)) + 
  geom_bar(stat="identity") + 
  labs(title="Absolute proportion of older planes by airline", x="Airline", y="Number of planes")
  
```
I categorized "old planes" as planes that are from before 1995. Based on that assumption, the airline with the highest absolute number of flights on older planes (from Houston) is by far WN (Southwest), followed (not closely) by CO (Continental) and US (US Air). However, proportionally, WN comes in 6th with only 33.4% of its flights on older planes. Proportionally, then, the top 4 airlines to fly with are MQ (Envoy), AA (American Airlines), US (US Air), and DL (Delta), bearing in mind that MQ, for example, only has an absolute number of 57 flights on older planes.




## Question 3:

* What states did Southwest Airlines' **flight paths** tend to fly to?
* What states did Southwest Airlines' **flights** tend to fly to?

For example, Southwest Airlines Flight 60 to Dallas consists of a single flight
path, but since it flew 299 times in 2013, it would be counted as 299 flights.

```{r, echo=FALSE, message=FALSE, fig.width=12, fig.height=6}
flight_airports <- inner_join(flights, airports, by=c("dest"="iata"))
#bad code
#southwest_bad <- flight_airports %>% 
#  group_by(flight, state) %>% 
#  filter(carrier == "WN") %>% 
#  tally() %>% 
#  arrange()
#table(southwest_bad$state)

#more bad code
#southwest_paths <- flight_airports %>% 
#  group_by(flight, state) %>% 
#  filter(carrier == "WN") %>% 
#  tally() %>% 
#  group_by(state) %>% 
#  select(-n) %>% 
#  tally()
  
#paths
southwest_paths <- flight_airports %>% 
  group_by(flight, state) %>% 
  filter(carrier == "WN") %>% 
  tally() %>% 
  group_by(state) %>% 
  summarise(n=n()) %>% 
  arrange(desc(n)) %>% 
  knitr::kable()
southwest_paths
#destinations
southwest_destinations <- flight_airports %>% 
  group_by(state) %>% 
  filter(carrier == "WN") %>% 
  tally() %>% 
  arrange(desc(n)) %>% 
  knitr::kable()
southwest_destinations

```


Both flight paths and flights tended to fly to Texas, followed by Florida and Louisiana. In general, the frequency of flight paths and total flights follow the same order in frequency of destination.



## Question 4:

I want to know proportionately what regions (NE, south, west, midwest) each 
carrier flies to/from Houston in the month of July.  Consider the `month()`
function from the `lubridate` package.

```{r, echo=FALSE, message=FALSE, fig.width=12, fig.height=6}

region_flights <- inner_join(flights, airports, by=c("dest"="iata")) %>% 
  inner_join(states, by="state")

july_regions <- region_flights %>% 
  filter(month(date) == 7) %>% 
  group_by(carrier, region) %>% 
  tally() %>% 
  rename(total_in_region = n)
july_carrier <- region_flights %>% 
  filter(month(date) == 7) %>% 
  group_by(carrier) %>% 
  tally() %>% 
  rename(total_in_carrier = n)
july_total <- left_join(july_regions, july_carrier, by="carrier") %>% 
  mutate(prop = round(total_in_region/total_in_carrier, 3))
  
july_total %>% knitr::kable()

july_graph <- ggplot(july_total, aes(x=carrier, y=prop, fill=region)) + geom_bar(stat="identity")
july_graph
```
